Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-20 transmission-appendix-plots.tex
move half of plot to appendix

\begin{document}
\ifSubfilesClassLoaded{
    \setcounter{chapter}{6}
}

\chapter{Estimating transmission} \label{transmission}

In this chapter, I use the estimate of the duration of RT-PCR positivity from combining information from the ATACCC and CIS studies (see \cref{E-imperf-test:sec:conclusion}) to investigate the feasibility of estimating incidence and transmission of SARS-CoV-2 from only the CIS data (described in \cref{E-intro:sec:cis}).
I will show that the CIS data is sufficient, demonstrating the value of surveys of this type for pandemic surveillance.

I take two approaches, backcalculation and a mechanistic model, both previously discussed in \cref{E-inc-prev:sec:infection-process}.

First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
There are several attractive features about this approach in the context here.
The backcalculation approach makes no assumptions about how transmission occurs; the assumptions have the potential to introduce bias.
It can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

In \cref{SEIR}, I use a mechanistic compartmental model, within the modelling framework introduced in \cref{E-SEIR:sec:mechanistic-models}.
This approach enables estimating both incidence and components of transmission.
Estimating transmission in addition to incidence answers important epidemiological questions, \eg whether changes in transmission can be explained by changes in contact rates.
I develop a novel observation model to allow the transmission model to be fit to the CIS data.

Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.
I show they give similar values, providing reassurance of the robustness of the results to modelling assumptions.
Overall, there is a bias-variance trade-off between the models.
Incidence and transmission have not previously been estimated from only SARS-CoV-2 prevalence surveys before.

\section{Backcalculation} \label{backcalc}

In this section, I will apply the theory introduced in \cref{E-inc-prev} to estimate incidence.
I view the CIS as a series of daily, cross-sectional prevalence surveys.
This is the same approach as previously used to estimate prevalence from the CIS (see \cref{E-biology-data:sec:MRP}).

An important consideration in this section is the non-response bias.
I adopt the poststratification method introduced in \cref{E-biology-data:sec:MRP} to correct for this bias.
Therefore, incidence must be estimated in each stratum before being poststratified to the target population and subpopulations of interest.
The strata are defined by at least age group, region, ethnicity, and sex due to differential non-response in the survey (see \cref{E-intro:sec:cis}).

\subsection{Approaches} \label{backcalc:sec:approach}

In \cref{E-inc-prev:sec:observation-process}, I showed that a deterministic approximation to the relationship between true prevalence and incidence is a good approximation for estimating incidence.
This relationship between incidence proportion and prevalence means that once one is inferred, the other is calculated via a deterministic function.
This gives two possible approaches for estimating incidence from prevalence data.

% \subsubsection{Possible approaches}

The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
The basis of the statistical model, in a Bayesian framework, would be as follows.
\begin{align}
    \vec{Z} &\dist F_Z \\
    P_{i,t}  &= \sum_{k=0}^{d_\text{max}-1} Z_{i,t-k} S(k+1) \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $\vec{Z}$ is a vector such that $Z_{i,t}$ is the incidence proportion in stratum $i$ on day $t$, and has prior $F_Z$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\rho$ is the overdispersion parameter of the beta-binomial distribution with prior $F_\rho$; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
Fitting this model to a prevalence survey such as the CIS (\eg using MCMC) gives a posterior distribution of $\vec{Z}$, the target of inference.

The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.

Specifying $F_Z$ would depend on the application of the approach, but should consider the need to provide structure to stabilise the problem (see \cref{E-intro:sec:disease-model}).
In the context of the CIS, it needs to allow borrowing information across time and/or strata; this is required because some stratum and time combinations have few positives (fewer than five per day).

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{E-inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P}  &\dist F_P \\
    \rho &\dist F_\rho \\
    Y_{i,t} \mid \vec{P}, \rho &\dist \BB(n_{i,t}, P_{i,t}, \rho)
\end{align}
where $F_P$ is analogous to $F_Z$ in providing a joint prior for the prevalence in each stratum and day combination, and all other notation is as in the incidence-inferred approach.
The first step is to fit this model, giving a posterior distribution for $\vec{P}$.

The second step is to transform the posterior distribution for $\vec{P}$ to a posterior distribution for $\vec{Z}$.
A posterior sample for $\vec{P}$ can be transformed to a posterior sample for $\vec{Z}$ based on \cref{E-inc-prev:eq:forward-substitute}, see \cref{backcalc:sec:methods} for details.

\subsection{Desirable properties}

The incidence-inferred approach is more parsimonious in being a single step procedure.
However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach within the computational constraints of the SRS.

It will be convenient to define a variable equating to the posterior initially estimated.
Let $\pi_{i,t} = Z_{i,t}$ under the inference-inferred approach and $\pi_{i,t} = P_{i,t}$ is the prevalence.

The chosen approach should have the following properties.
\begin{enumerate}
    \item The prior should favour $\pi_{i,t}$ changing exponentially in $t$, \ie penalize changes from exponential growth.
        % Penalizing changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Such penalization has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately} and is the approach used for the CIS prevalence model (see \cref{E-biology-data:sec:MRP}).
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
    \item Borrow information across the different covariates.
        The borrowing should penalize changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

A parsimonious model satisfying properties 1 and 2 is to specify $\logit(\pi_{i,t}) = g_\varphi(i, t)$, where $g$ is a function parameterised by $\varphi$ and the prior on $\varphi$ is specified to favour $d^2g/dt^2 = 0$, \ie favour linear growth in $\logit(\pi_{i,t}$.
The prior $F_Z$ or $F_P$ is now a prior on $\varphi$.
The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.

The logit transformation also allows property 4 to be conveniently satisfied using an additive model for $g_\varphi$.
An additive model on the logit scale is multiplicative on the natural scale for the small values of $\pi_{i,t}$ which will be common.

This form of model includes the MRP prevalence model in \cref{E-biology-data:sec:MRP}.

% Combined with property 4, the form of $g$ the null model is $g_\varphi(i, t) = \varphi_0 + \varphi_1 t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalized.

% Properties 3 and 4 suggest the extension of the previously-developed MRP method, used to already estimate prevalence, which already satisfies these properties.
% This method has previously been used successfully to estimate prevalence, corrected for non-response bias, from the CIS data~\autocite{cisMethodsONS,pouwelsCommunity,pouwelsMRPvaccination}.

Properties 3 and 5 suggest the use of an approximate method for computing the posterior distribution.
In particular, using MCMC, sufficient covariates to adjust for the non-response bias, and a period of more than 6--8 weeks leads to a model that is computationally prohibitive ~\citePersonalComms{Koen Pouwels}.
Instead, previous MRP implementations used INLA (see \cref{E-biology-data:sec:MRP}).

INLA-based inference can only be applied to LGMs (\emph{Latent Gaussian Models}); see \cref{E-transmission:sec:INLA} for further details on INLA and LGMs.
The LGM framework requires that a transformation of $P_{i,t}$ is linear in $g_\varphi = \logit(\phi_{i,t})$.

If $\pi_{i,t} = P_{i,t}$ then this is trivially satisfied.
However, if $\pi_{i,t} = Z_{i,t}$, then no such linear relationship exists.
In particular:
\begin{align}
    P_{i,t}
    &= \sum_{k=0}^{\dmax} \pi_{i,t-k} S(k+1) \\
    &= \sum_{k=0}^{\dmax} \expit(g_\varphi(i, t-k)) S(k+1).
\end{align}

Therefore, satisfying all the desirable properties requires the prevalence-inferred approach, which I adopt.

\todo[inline]{Here to the end of the section moved to Appendix.}

An LGM is defined in terms of a vector of hyperparameters, $\vec{\varphi}$; a deterministic transformation of the hyperparameters into covariance matrix, $\matr{\Sigma}(\vec{\varphi})$; a vector of latent variables, $\vec{z}$; a vector of $n$ data points, $\vec{w} = [w_1, \dots, w_n]^T$; and $n$ linear predictors $\vec{\eta} = [\eta_1, \dots, \eta_n]^T$.
To be an LGM, $\vec{z}$ must be a latent Gaussian field, that is $\vec{z} \mid \vec{\varphi} \dist \N(\vec{0}, \matr{\Sigma}(\vec{\varphi}))$; and the likelihood must have the conditional independence property $p(\vec{w} \mid \vec{\varphi}, \vec{z}) = \prod_j p(w_j \mid \eta_j(\vec{z}), \vec{\varphi})$.

In this context, each $w_j$ corresponds to the observed RT-PCR results in one stratum on one day.
Hence, $\eta_{j}$ must determine the distribution of $w_{j}$ (up to some hyperparameters); where $j$ uniquely identifies a stratum $i$ and day $t$.

In practice, this means that $\eta_j$ is a one-to-one transformation of the population prevalence in stratum $i$ on day $t$.
If $\eta_j = \logit(\pi_{i,t})$, where $\pi$ is prevalence, then a model of $\eta_j = g(i, t)$, as before, is an LGM.
However, if $\pi$ is instead representing incidence proportion, then no such linear relationship between $\eta_j$ and $g$ exists.
In particular, if $\eta_j$ is the prevalence, then:
\begin{align}
    \eta_{i,t}
    &= \sum_{k=-\infty}^0 \pi_{ik} S(k) \\
    &= \sum_{k=-\infty}^0 \expit(g(i, k)) S(k).
\end{align}
As $g(i, k)$ is linear in the model parameters, $\vec{z}$, then $\eta_{i,t}$ is not, violating the requirements of an LGM.

\subsection{Methods} \label{backcalc:sec:methods}

Now I explain how I extended the MRP model introduced in \cref{E-biology-data:sec:MRP} to implement the prevalence-inferred approach.
I start with the prevalence posterior from \cref{E-biology-data:sec:positivity-results}, denoting the times the estimates are for as $t = 1, \dots, T$.
I draw 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).

Then, I transform this posterior sample using \cref{transmission:alg:calculate-Z}.
The algorithm is based on \cref{E-inc-prev:eq:forward-substitute}, which is repeated here for convenience:
\begin{align}
Z_t
% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
&= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
\label{transmission:eq:forward-substitute}
\end{align}
Dividing each side of \cref{transmission:eq:forward-substitute} by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.
The algorithm uses a deterministic relationship between incidence and prevalence within a stratum, as \cref{E-inc-prev:sec:observation-process} justifies; \ie, the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.

\begin{algorithm}
    \For{$j = 1, \dots, 500$}{
      Set $\vec{P}^{(j)}$ to a posterior sample of $\vec{P}$ \;
      \For{each stratum $i$}{
        \For{$t = -\dmax, \dots, 0$}{
            $Z^{(j)}_{i,t} = P^{(j)}_{i,1} / \sum_{k=1}^{\dmax} S(k)$
        }
        \For{$t = 1, \dots, T$}{
            Set $Z^{(j)}_{i,t} = P^{(j)}_{i,t} - \sum_{k=-\dmax}^{t-1} S(t - k + 1) Z^{(j)}_{i,k}$ \;
        }
      }
    }
    \caption{%
      Algorithm to transform 500 posterior samples posterior samples of $\vec{P}$ to posterior samples of $\vec{Z}$.
      Assumes constant incidence for $t \leq 1$ where $t = 1$ is the first day of prevalence data.
      Calculation of $Z^{(j)}_{i,t}$ is from \cref{transmission:eq:forward-substitute}.
    }
    \label{transmission:alg:calculate-Z}
\end{algorithm}

\Cref{transmission:eq:forward-substitute} is derived from \cref{E-inc-prev:eq:EPt-matrix}: $\vec{P} = \matr{S} \vec{I}$.
For this to have a unique solution, $\vec{I}$ and $\vec{P}$ need the same dimension.
However, I am using a period of prevalence which starts after the first infections.
These prior infections will still contribute to the prevalence.
I assume that the incidence for $t \leq 1$ is constant.
Therefore, $P_1 = I_0 \sum_{i=1}^{\dmax} S(i)$ and $I_t = P_1 / \sum_{i=1}^{\dmax} S(i)$ for $t \leq 1$.

% First, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution (using the procedure described in \cref{transmission:sec:INLA:posterior}).
% For each sample and stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
% I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:observation-process} justifies, that is the estimate of the incidence is negligibly affected by approximating the distribution $\vec{P} \mid \vec{Z}$ by its mean.
% I solve this system using forward substitution, calculating the value of $Z_t$ in \cref{E-inc-prev:eq:forward-substitute} sequentially for $t = 1, \dots, T$.
% $t = 1$ is defined as the first day where prevalence data is used.
% \Cref{E-inc-prev:eq:forward-substitute} is repeated here for convenience:
% \begin{align}
% Z_t
% &= \frac{P_t - \sum_{i=1}^{t-1} S_{ti} Z_i}{S_{tt}} \\
% &= P_t - \sum_{i=1}^{t-1} S(t - i + 1) Z_i.
% \label{transmission:eq:forward-substitute}
% \end{align}
% % \Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
% Dividing each side by the total number of individuals in each stratum transforms the number of prevalent individuals to prevalence and incidence to incidence proportion.

Violation of assumptions in various parts of this procedure can lead to the second term in \cref{transmission:eq:forward-substitute} to be greater than the first, giving negative incidence.
In this case, the incidence prior to day $t$ imply that the prevalence on day $t$ is higher than that predicted by the fitted regression model for this strata/day combination.
There are two sets of assumptions which are most likely to be violated.

The first set of assumptions is in estimating the survival function, $S(t)$, which is assumed to be known with no uncertainty in this procedure.
However, \cref{E-imperf-test} showed that the estimate of $S(t)$ is sensitive to assumptions made, such as the assumed $\psens$ and strength of priors used.

The second set of assumptions are in specifying the prior for $\vec{P}$.
In particular, the prior places non-zero mass on prevalence values which imply negative incidence under the other model assumptions.
Rectifying this assumption is challenging without switching to the inference-inferred approach, discussed further in \cref{backcalc:sec:results}.

For each posterior sample of $\vec{Z}$, I poststratify it to the CIS target population and subpopulations of interest.
This is done for each posterior sample using the methodology explained in \cref{E-biology-data:sec:MRP}.
The poststratified posterior samples are now posterior samples from the posterior of interest.

\subsection{Results and discussion} \label{backcalc:sec:results}

A slow increase in incidence is estimated to occur over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
All regions then decrease during November; this coincides with the national lockdown which came into force in England on 5th Nov 2020~\autocite{ifgTimeline}.

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for (A) all of England and (B) its component regions.
        Yorkshire and the Humber is labelled Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
    \label{transmission:fig:backcalc-alpha}
\end{figure}

The incidence proportion was highest among the 11--15 and 16--25 age groups throughout the period (see \cref{transmission:fig:backcalc-ages}).
Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale), meaning there was insufficient information in the data to overcome the prior which favoured this situation.
The incidence proportion in the 16--25 age group decreases faster than other age groups during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means it could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown; 20--24 year olds rarely interact with school-aged children (see \cref{E-SEIR:fig:age-contacts})
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using backcalculation by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

An edge effect where a rapid rise is seen between day 1 and 2 is visible (see \cref{transmission:fig:backcalc-start-effect}).
This is due to the incorrect assumption that incidence is constant in the past.
Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
Therefore, the first day's estimate is unreliable, and has been excluded from all other plots.
However, the rest of the series is robust to this assumption and hence reliable.
\begin{figure}
    \centering \includegraphics{transmission/backcalc-start-effect}
    \caption[Edge effects in backcalculation method]{%
        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
    }
    \label{transmission:fig:backcalc-start-effect}
\end{figure}

Alternative inference techniques could allow more powerful modelling approaches, if they allowed a broader set of models than LGMs.
In particular, the incidence-inferred approach could be adopted.
This could allow more interpretable priors to be used because the infection process is the process that corresponds to biological and behavioural processes.
Furthermore, the incidence proportion could be restricted to be non-negative, preventing the negative incidence seen in some strata.

Various alternative inference techniques could be considered.
If the infection process is considered a Poisson process (defined in \cref{E-inc-prev:sec:poisson-process}), then, conditional on the process's intensity, there is independence between individuals' infection status.
The conditional independence implies the incidence proportion and prevalence are the mean of many independent Bernoulli variables.
Therefore, the Central Limit Theorem implies they are approximately normally distributed.
Exploiting this property could enable efficient approximations to the posterior.
For example, a Laplace approximation or variational inference~\autocite{bleiVariational}.
These methods require further work to check the quality of the approximations in this context.

\section{Mechanistic modelling} \label{SEIR}

In this section, I apply the general mechanistic transmission modelling framework from \cref{E-SEIR} to the CIS prevalence data.
The application of this framework follows \textcite{birrellRealtime}.
The largest modification from the generic framework presented in \cref{E-SEIR} is the parameterization of the WAIFW matrix (\cref{SEIR:sec:WAIFW-application}).
A novel approach (\cref{SEIR:sec:observation}) links this transmission model to the CIS prevalence data; the approach is generalisable to other studies collecting representative prevalence data, \eg REACT.
In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, in the terminology of \cref{E-inc-prev}, it includes both the prevalence process and the observation process.
Having specified the full model (\cref{SEIR:sec:methods-application}), I conducted a simulation study to validate the inference process and identifiability (\cref{SEIR:sec:sim-study}), before applying the model to the CIS data (\cref{SEIR:sec:application}).
In \cref{SEIR:sec:discussion}, I discuss the results; in \cref{transmission:sec:comparison} I compare the results to those in the previous section.

In common with \cref{E-SEIR:sec:mechanistic-models}, the notation in this section differs slightly from the rest of the thesis.

\subsection{Methods} \label{SEIR:sec:methods-application}

% The model will be stratified only by region and age group, so the data are aggregated to these levels.
% Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.

I model each region fully independently, \ie with no shared parameters or transmission between regions.
As well as the motivation given in \cref{E-backcalc}, independent regions allow heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
By modelling each region independently, I assume that the epidemics in each region do not interact.
Clearly, this is not realistic as travel between regions of England is common and unrestricted.
However, except in an epidemic's early stages, the interaction between large regions has a negligible impact on the epidemic's progression~\autocite[e.g.][]{birrellRealtimea,gogSpatial,eggoSpatial}.

\subsubsection{WAIFW matrix} \label{SEIR:sec:WAIFW-application}

A parsimonious model is needed to parameterize the WAIFW matrix.
In \cref{E-SEIR:sec:structured-populations}, I explained how the elements of the WAIFW matrix, $\matr{\beta}$ can be decomposed as $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'} / N_{a}$ where\todo{should I repeat these definitions? they already exist in ch 3}: $\kappa_{aa'}$ is the mean rate at which an individual in stratum $a'$ makes contact with individuals in stratum $a$; $q_{aa'}$ is the probability of transmission upon contact between an infectious individual in stratum $a'$ and a susceptible individual in stratum $a$; and $N_{a}$ is the total number of individuals in stratum $a$.
For many epidemics, where the population behaviour is fairly constant over time, a constant $\matr{\beta}$ is a good approximation.
However, in SARS-CoV-2, population behaviour changed dramatically, both spontaneously (\eg due to fear) and in response to government-imposed restrictions.
Therefore, following \textcite{birrellRealtime}, I introduce a time-dependence giving: $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'}(t) N_{a'} / N_a$; $N_a$ and $N_{a'}$ are constant because the change in the total population is negligible over the length of time considered here.

I take information on the contact rates, $\kappa_{aa'}(t)$, from several data sources, using the methodology of \textcite{vanleeuwenTime,vanleeuwenAugmenting}.
%; these estimates have been used for previous modelling of the SARS-CoV-2 pandemic~\autocite{vanleeuwenTime,birrellRealtime}.
% These start with estimates of contact in each setting, based on \textcite{vanleeuwenAugmenting}.
This method starts with pre-pandemic baseline estimates of $\kappa_{aa'}$ from the POLYMOD study~\autocite{mossongSocial}, and time use from the UK Time Use Survey (UKTUS)~\autocite{UKTUS}.
POLYMOD asked a representative sample of the UK population about the contacts they make each day, including information on each contact's age and location (home, work, school, leisure, transport, or other).
UKTUS asked a representative sample of the UK population about how much time they spent doing various activities (\eg shopping and spending time at home) each day.
\Textcite{vanleeuwenAugmenting} combined these two datasets to estimate the rate of contacts between age groups during each activity.
\Textcite{birrellRealtime} then modify these contact rates based on measured changes in time use during the pandemic.
These changes are measured using Google mobility data~\autocite{googleCOVID19} and school attendance data~(unpublished).
Google mobility data are based on the location of Android users over time, relative to a pre-pandemic baseline.
The mobility and school data are averaged over a week then used to down or up-weight the number of contacts in each setting, before combining contacts across settings to form $\kappa_{aa'}(t)$.
$\kappa_{aa'}(t)$ is assumed to be piecewise constant, changing only at the start of each week.
For further details see \textcite{vanleeuwenAugmenting} and \textcite[supplementary material]{birrellRealtime}.

The probability of transmission, $q_{aa'}(t)$, needs further decomposition to be identifiable, for the same issue of dimensionality preventing each element of $\matr{\beta}$ being estimated independently (see \cref{E-SEIR:sec:structured-populations} for a full explanation).
I assume that $q_{aa'}(t) = \beta(t) q^\text{susc}_{a} q^\text{inf}_{a'}$ where $\beta(t)$ is a time-varying reference probability of transmission; $q^\text{susc}_a$ is the relative \emph{susceptibility} of stratum $a$, \ie how vulnerable to infection individuals in stratum $a$ are; and $q^\text{inf}_{a'}$ is the relative \emph{transmissibility} of stratum $a'$, \ie how likely individuals in stratum $a$ are to pass on an infection.
These measures include both biological differences, \eg due to differences in immune systems, or behavioural, \eg if infants are less hygienic than adults.

The time-varying component, $\beta(t)$, allows for behavioural changes that impact transmission but are not captured by the contact matrices (\eg mask wearing).
I impose a random walk prior on $\log\beta(t)$ on the weekly timescale, smoothing its value over time.
This prior encodes that behaviour changes smoothly over time.
\begin{align}
    \beta(t) &= \beta_{\tilde{w}(t)} \\
    \log\beta_{w+1} &= \log\beta_w + \sigma_\epsilon \epsilon_w \\
    \epsilon_w &\sim \N(0, 1)
\end{align}
where $\tilde{w}(t)$ is the week number of time $t$, with week 1 being the first week considered.
Two hyperparameters will be inferred: $\beta_1$, the initial value of the random walk; and $\sigma_\epsilon$, the standard deviation of the random walk (see \cref{SEIR:sec:priors}).
$\beta(t)$ is the only time-varying part of the transmission model that is not fixed a priori.
Therefore, in practice, it will also absorb other effects such as any error in $\matr{\kappa}$.

% $K_{ij}$ and $\beta_{ij}$ appear only as a product and therefore are not individually identifiable.
% Normally, $K_{ij}$ is assumed to be known, for example from a previous contact survey~\autocite[such as][]{mossongSocial}.
% Ideally, all the $\beta_{ij}$ terms would be estimated, but there is normally inadequate data to do so, requiring assumptions to be made to reduce the number of parameters~\autocite[65]{keelingModeling}.
% % $\beta_{ij}$ is then estimated, although often with a parsimonious model
% A common assumption is that $\beta_{ij}$ can be decomposed into two independent components such that $\beta_{ij} = \beta \beta^\text{susc}_{i} \beta^\text{inf}_j$.
% For any stratum $i$, $\beta^\text{susc}_i$ is known as the \emph{relative susceptibility} of stratum $i$.

%Now, we assume that $\beta_{ij} = \beta(t) \beta_{ij}$ where $\beta(t)$ is a non-strata-specific, time-varying component representing precautions taken by the population, for example masking; and $\beta_{ij}$ represents a constant probability of transmission from an infectious individual in strata $j$ to a susceptible individual in strata $i$ given contact, which could, for instance, be affected by differential susceptibility to infection for different age-groups.
%We arbitrarily choose one $\beta_{ij}$ to be equal to 1 for identifiability and hence the other $\beta_{ij}$ values are relative to this reference class.

Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocite{chenRole,vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for 0--11 year olds, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ can be interpreted as the relative probability that a child is infected (compared to an adult), given contact with an infectious individual.
%$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

\subsubsection{Prevalence and observation models} \label{SEIR:sec:observation}

This section  explains the novel observation model approach I propose for fitting to the CIS prevalence data.
This will link the model in the previous section to the CIS prevalence data.

% Here, I extend the system of ODEs to include compartments representing RT-PCR positive individuals.
% The proportion of RT-PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are RT-PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the RT-PCR duration distribution.
Here, the daily incidence proportion for each stratum derived by solving the system of ODEs governing the transmission model (see \cref{E-SEIR:sec:SIR}), and then prevalence calculated using \cref{E-inc-prev:eq:EPt}.
However, this approach is computationally expensive, likely prohibitively expensive if performed within a MCMC loop.

A computationally cheaper option, which I will take, is to approximate this convolution by adding compartments to the ODEs.
This approach allows an infected, pre-positive compartment, which I denote $p_0$, to be included.
Up until now, this stage of the disease (shown in \cref{E-biology-data:fig:natural-history}) has been neglected, assumed to be of negligible duration.
An advantage of this modelling approach is that it can be included without complications (\eg issues of numerical stability).

$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartments, $p_1, \dots, p_l$, each with their own transition rate.
I will choose $l$ and the transition rates such that combined length of stay (\ie time from entry in $p_1$ until exit from $p_l$) approximates the duration of positivity distribution estimated in \cref{E-imperf-test}.
This idea extends the technique of having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).
While the use of exponential or Erlang (gamma distributions with integer first parameter) is common in the observation model of compartmental models~\autocite[e.g.][]{overtonEpiBeds}, the use of more flexible phase-type distributions, as here, has rarely, if ever, been used in practice~\autocite{hurtadoGLCT}.
This broader class of distributions allows the model to capture the tail of the distribution, which is important when incidence suddenly decreases.

\Textcite{osogamiClosed} provide a method to construct the compartments, implemented in the mapfit R package~\autocite{mapfit}. 
The resulting compartmental approximation matches the original distribution up to the first three moments.
Applying the package to the posterior mean of the distribution estimated in \cref{E-imperf-test} gives the model in \cref{SEIR:fig:full-model} with the parameter estimates in \cref{SEIR:table:ec-params} (see \cref{transmission:sec:phase-type} for details).
\begin{table}
    \centering
    \begin{tabular}{c c c c c}
        $l$ & $\omega_1$ & $\omega_{2}$ & $\omega_{3}$ & $\pi$ \\
        3 & 0.0820 & 0.126 & 0.0223 & 0.00973  \\
    \end{tabular}
    \caption{Parameter estimates for the EC distribution approximating the posterior mean of the duration distribution estimated in \cref{E-imperf-test}.}
    \label{SEIR:table:ec-params}
\end{table}
% end table of ec params

I link the modelled proportion of individuals that are RT-PCR positive to the data using a beta-binomial likelihood (as discussed in \cref{E-biology-data:sec:clustering}).
If on each day $t$, each stratum $i$ is observed to have $y_{at}$ positives out of $n_{at}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \bar{p}_{at}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where: $\bar{p}_{at}$ is the mean proportion of individuals in strata $a$ that are RT-PCR positive on day $t$; and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
Denoting the proportion of stratum $a$ in state $p_k$ as $p_{ka}(\tau)$, then $\bar{p}_{at} = \int_{t}^{t+1} \sum_{k=1}^l p_{ka}(\tau) d\tau$.

\subsubsection{Model equations and solution} \label{SEIR:sec:full-model}

>>>>

\begin{document}

\chapter{Additional plots for \cref{E-transmission}}

\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_appendix}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[Other SEIR prevalence goodness-of-fit]{%
        Goodness-of-fit to prevalence data by region and age for selected regions (see \cref{SEIR:fig:prev} for others).
        The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev-appendix}
\end{figure}
